Testing gene methylation enrichment analysis approaches using simulated data

### 2023-12-21 figs/GSE158422_simulate050.Rmd
fix gene catalog

simalm <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
  # generate gene sets
  gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))
  lengths <- unname(unlist(lapply(genesetdatabase,length)))
  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
  # select gene sets to alter
  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
  # incorporate select changes
  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta,
    gene_catalog=gene_catalog)
  # set up limma
  mval2 <- sim$mval
  ncols <- ncol(mval2)
  groupsize <- ncols/2
  ss <- data.frame(colnames(mval2))
  colnames(ss) <- "sample"
  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
  d <- model.matrix(~ ss$case )
  # alm pipeline
  mm <- merge(mval2,gt,by.x=0,by.y="probe")
  mm$Row.names = NULL
  a <- aggregate(. ~ gene,mm,mean)
  rownames(a) <- a$gene
  a$gene=NULL
  fit.reduced <- lmFit(a,d)
  fit.reduced <- eBayes(fit.reduced)
  al <- topTable(fit.reduced,coef=ncol(d), number = Inf)
  m1 <- as.data.frame(al$t)
  rownames(m1) <- rownames(al)
  colnames(m1) <- "t"
  almres1 <- runmitch(m=m1,genesets=gsets,cores=4)
  # summarise results
  gsig_up3 <- rownames( subset( almres1, p.adjustANOVA < 0.05 & s.dist > 0 ) )
  gsig_dn3 <- rownames( subset( almres1, p.adjustANOVA < 0.05 & s.dist < 0 ) )
  gtup <- names(sim[[6]])
  gtdn <- names(sim[[7]])
  UPTP=length(intersect(gsig_up3 ,gtup))
  UPFP=length(setdiff(gsig_up3 ,gtup))
  UPFN=length(setdiff(gtup,gsig_up3))
  DNTP=length(intersect(gsig_dn3 ,gtdn))
  DNFP=length(setdiff(gsig_dn3 ,gtdn))
  DNFN=length(setdiff(gtdn,gsig_dn3))
  TP=UPTP+DNTP
  FP=UPFP+DNFP
  FN=UPFN+DNFN
  TN=nrow(almres1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
  PREC=TP/(TP+FP)
  REC=TP/(TP+FN)
  F1=TP/(TP+(0.5*(FP+FN)))
  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
  return(result)
}

@@

simaa <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
  # generate gene sets
  gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))
  lengths <- unname(unlist(lapply(genesetdatabase,length)))
  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
  # select gene sets to alter
  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
  # incorporate select changes
  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta,
    gene_catalog=gene_catalog)
  # set up limma
  mval2 <- sim$mval
  ncols <- ncol(mval2)
  groupsize <- ncols/2
  ss <- data.frame(colnames(mval2))
  colnames(ss) <- "sample"
  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
  d <- model.matrix(~ ss$case )
  mm <- merge(mval2,gt,by.x=0,by.y="probe")
  mm$Row.names = NULL
  a <- aggregate(. ~ gene,mm,mean)
  rownames(a) <- a$gene
  a$gene=NULL
  mystack <- stack(gsets)
  mmm <- merge(a,mystack,by.x=0,by.y="values")
  mmm$Row.names=NULL
  aa <- aggregate(. ~ ind,mmm,mean)
  rownames(aa) <- aa$ind
  aa$ind=NULL
  fit.reduced <- lmFit(aa,d)
  fit.reduced <- eBayes(fit.reduced)
  aares1 <- topTable(fit.reduced,coef=ncol(d), number = Inf)
  # summarise results
  gsig_up3 <- rownames(subset(aares1, adj.P.Val < 0.05 & logFC > 0))
  gsig_dn3 <- rownames(subset(aares1, adj.P.Val < 0.05 & logFC < 0))
  gtup <- names(sim[[6]])
  gtdn <- names(sim[[7]])
  UPTP=length(intersect(gsig_up3 ,gtup))
  UPFP=length(setdiff(gsig_up3 ,gtup))
  UPFN=length(setdiff(gtup,gsig_up3))
  DNTP=length(intersect(gsig_dn3 ,gtdn))
  DNFP=length(setdiff(gsig_dn3 ,gtdn))
  DNFN=length(setdiff(gtdn,gsig_dn3))
  TP=UPTP+DNTP
  FP=UPFP+DNFP
  FN=UPFN+DNFN
  TN=nrow(aares1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
  PREC=TP/(TP+FP)
  REC=TP/(TP+FN)
  F1=TP/(TP+(0.5*(FP+FN)))
  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
  return(result)
}

>>>>

simalm <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
  # generate gene sets
  gene_catalog <- unique(gt$gene)
  lengths <- unname(unlist(lapply(genesetdatabase,length)))
  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
  # select gene sets to alter
  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
  # incorporate select changes
  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta,
    gene_catalog=gene_catalog)
  # set up limma
  mval2 <- sim$mval
  ncols <- ncol(mval2)
  groupsize <- ncols/2
  ss <- data.frame(colnames(mval2))
  colnames(ss) <- "sample"
  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
  d <- model.matrix(~ ss$case )
  # alm pipeline
  mm <- merge(mval2,gt,by.x=0,by.y="probe")
  mm$Row.names = NULL
  a <- aggregate(. ~ gene,mm,mean)
  rownames(a) <- a$gene
  a$gene=NULL
  fit.reduced <- lmFit(a,d)
  fit.reduced <- eBayes(fit.reduced)
  al <- topTable(fit.reduced,coef=ncol(d), number = Inf)
  m1 <- as.data.frame(al$t)
  rownames(m1) <- rownames(al)
  colnames(m1) <- "t"
  almres1 <- runmitch(m=m1,genesets=gsets,cores=4)
  # summarise results
  gsig_up3 <- rownames( subset( almres1, p.adjustANOVA < 0.05 & s.dist > 0 ) )
  gsig_dn3 <- rownames( subset( almres1, p.adjustANOVA < 0.05 & s.dist < 0 ) )
  gtup <- names(sim[[6]])
  gtdn <- names(sim[[7]])
  UPTP=length(intersect(gsig_up3 ,gtup))
  UPFP=length(setdiff(gsig_up3 ,gtup))
  UPFN=length(setdiff(gtup,gsig_up3))
  DNTP=length(intersect(gsig_dn3 ,gtdn))
  DNFP=length(setdiff(gsig_dn3 ,gtdn))
  DNFN=length(setdiff(gtdn,gsig_dn3))
  TP=UPTP+DNTP
  FP=UPFP+DNFP
  FN=UPFN+DNFN
  TN=nrow(almres1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
  PREC=TP/(TP+FP)
  REC=TP/(TP+FN)
  F1=TP/(TP+(0.5*(FP+FN)))
  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
  return(result)
}

@@

simaa <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
  # generate gene sets
  gene_catalog <- unique(gt$gene)
  lengths <- unname(unlist(lapply(genesetdatabase,length)))
  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
  # select gene sets to alter
  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
  # incorporate select changes
  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta,
    gene_catalog=gene_catalog)
  # set up limma
  mval2 <- sim$mval
  ncols <- ncol(mval2)
  groupsize <- ncols/2
  ss <- data.frame(colnames(mval2))
  colnames(ss) <- "sample"
  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
  d <- model.matrix(~ ss$case )
  mm <- merge(mval2,gt,by.x=0,by.y="probe")
  mm$Row.names = NULL
  a <- aggregate(. ~ gene,mm,mean)
  rownames(a) <- a$gene
  a$gene=NULL
  mystack <- stack(gsets)
  mmm <- merge(a,mystack,by.x=0,by.y="values")
  mmm$Row.names=NULL
  aa <- aggregate(. ~ ind,mmm,mean)
  rownames(aa) <- aa$ind
  aa$ind=NULL
  fit.reduced <- lmFit(aa,d)
  fit.reduced <- eBayes(fit.reduced)
  aares1 <- topTable(fit.reduced,coef=ncol(d), number = Inf)
  # summarise results
  gsig_up3 <- rownames(subset(aares1, adj.P.Val < 0.05 & logFC > 0))
  gsig_dn3 <- rownames(subset(aares1, adj.P.Val < 0.05 & logFC < 0))
  gtup <- names(sim[[6]])
  gtdn <- names(sim[[7]])
  UPTP=length(intersect(gsig_up3 ,gtup))
  UPFP=length(setdiff(gsig_up3 ,gtup))
  UPFN=length(setdiff(gtup,gsig_up3))
  DNTP=length(intersect(gsig_dn3 ,gtdn))
  DNFP=length(setdiff(gsig_dn3 ,gtdn))
  DNFN=length(setdiff(gtdn,gsig_dn3))
  TP=UPTP+DNTP
  FP=UPFP+DNFP
  FN=UPFN+DNFN
  TN=nrow(aares1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
  PREC=TP/(TP+FP)
  REC=TP/(TP+FN)
  F1=TP/(TP+(0.5*(FP+FN)))
  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
  return(result)
}
